---
title: "Statistical inference with the General Social Survey data"
output: 
  html_document: 
    fig_height: 3
    highlight: pygments
    theme: spacelab
---

### Setup

#### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(graphics)
library(corrplot)
```

#### Load data

```{r load-data}
load("gss.Rdata")
```

* * *

### Part 1: Data

The [General Social Survey](https://www.norc.org/Research/Projects/Pages/general-social-survey.aspx) (GSS) is a sociological survey created and regularly collected since 1972 by the National Opinion Research Center at the University of Chicago. It is funded by the National Science Foundation. The GSS collects information and keeps a historical record of the concerns, experiences, attitudes, and practices of residents of the United States.

The target population of the GSS is adults (18+) living in households in the United States. The GSS sample is drawn using an area probability design that randomly selects respondents in households across the United States to take part in the survey. Respondents that become part of the GSS sample are from a mix of urban, suburban, and rural geographic areas. Participation in the study is strictly voluntary.

The survey is conducted face-to-face with an in-person interview by NORC at the University of Chicago. Computer-assisted personal interviewing (CAPI) began in the 2002 GSS. The survey was conducted every year from 1972 to 1994 (except in 1979, 1981, and 1992). Since 1994, it has been conducted every other year. As of 2014, 30 national samples with 59,599 respondents and 5,900+ variables have been collected. [1]

##### Generalisability

The sample from the GSS should be representative of the target population as respondents were randomly selected. Methodology used for the sample collection was designed to neutralise the possible sources of bias, specifically "voluntary" bias. Researches weighted observations. 

##### Causality

As the GSS is an observational study with no random assignment, any causal relationships of the variables cannot be inferred.

* * *

### Part 2: Research questions

Abortions have always been an acute and controversial topic. Considering the recent news on the near-total abortion ban in Poland and Texas, it is probable that more changes regarding women's rights are to be expected. I would like to look into the structure of the public opinion on abortions: namely, whether there is a correlation between respondents' opinions and their education level and also their age groups. 
Respondents were asked whether or not they think it should be possible for a pregnant woman to obtain a legal abortion if the woman wants it for any reason. 

##### 2.1. Is there a correlation between respondent's education level and their opinion on the topic and which education level groups in population have stronger opinions?

Variables used: 
degree - respondent's highest degree (categorical with 5 levels):
  Less than high school
  High school
  Associate/junior college
  Bachelor's
	Graduate

abany - respondent's opinion on abortion if woman wants for any reason (categorical with 2 levels).

##### 2.2. Is there a correlation between respondent's age group and their opinion on the topic and which age groups in population have stronger opinions?

Variables used: age - respondent's age (numerical),
abany - respondent's opinion on abortion if woman wants for any reason (categorical with 2 levels).

##### 2.3. Is there a correlation between respondent's gender and their opinion on the topic?

Variables used: respondent's sex (categorical with 2 levels),
abany - respondent's opinion on abortion if woman wants for any reason (categorical with 2 levels).

* * *

### Part 3: Exploratory analysis

#### 3.1. Education level group and opinion on abortions

Let's see the number of available observations, that have all variables of interest presented, by year and by education level groups:

```{r}
abor_edu <- select(gss, year, abany, degree, sex) %>%
  # Filter out NAs
  filter(is.na(abany) == FALSE, is.na(degree) == FALSE, is.na(sex) == FALSE)

table(abor_edu$year)
table(abor_edu$degree)
```

Contingency table for the variables of interest:

```{r}
cont_t_edu <- table(abor_edu$abany, abor_edu$degree)
cont_t_test_edu <- table(abor_edu$degree, abor_edu$degree)

cont_t_edu
```

Let's look into the most recent data available in our dataset which was the year of 2012:

```{r}
abor_edu_2012 <- abor_edu[abor_edu$year == 2012,]

table(abor_edu_2012$degree)
table(abor_edu_2012$degree,abor_edu_2012$abany)
```

The graph of distribution of proportion of Yes and No answers among education levels groups shows a steady decrease of No answers with the growth of respondent's level of education:

```{r}
ggplot(abor_edu_2012) +
  aes(x = degree, fill = abany) +
  geom_bar(position = "fill")
```

Contingency table for 2012:

```{r}
cont_t_edu_2012 <- table(abor_edu_2012$abany, abor_edu_2012$degree)
cont_t_edu_2012_m <- table( abor_edu_2012$degree, abor_edu_2012$abany)
cont_t_edu_2012_m
```

Let's see which degree group gave the biggest contribution in the observed difference in opinions:

```{r}
mosaicplot(cont_t_edu_2012_m, shade = TRUE, las=2,
           main = "degree")
```

Blue color in this mosaic plot indicates that the observed value is higher than the expected value if the data were random. Red color specifies that the observed value is lower than the expected value if the data were random.

#### 3.2. Age group and opinion on abortions 

```{r}
abor_age <- select(gss, year, abany, age, sex) %>%
  # Filter out NAs
  filter(is.na(abany) == FALSE, is.na(age) == FALSE, is.na(sex) == FALSE) %>%
  group_by(sex) 

#the latest year available in the data is 2012: 
abor_age_2012 <- abor_age[abor_age$year == 2012,]
```

Let's split variable age into age groups so that each group would contain close number of observations:

```{r}
abor_age_2012["age_group"] = cut(abor_age_2012$age, c(18, 25, 30, 35, 40, 45, 50, 55, 60, 70, Inf), c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-69", ">70"), include.lowest=TRUE)

table(abor_age_2012$age_group)
```

As the graph of proportions of Yes answers for different year groups indicates, people in age group of 35-49 seem to have the highest rate of Yes answers, which might be due to them be so-called “Sandwich Generation" usually carry the burden of looking after children and older relatives. The lowest rate of Yes answers was for people in age group > 70:

```{r}
ggplot(abor_age_2012) +
  aes(x = age_group, fill = abany) +
  geom_bar(position = "fill")
```

Contingency table for the variables of interest:

```{r}
cont_t_2012 <- table( abor_age_2012$abany, abor_age_2012$age_group)
cont_t_2012_m <- table( abor_age_2012$age_group, abor_age_2012$abany)
cont_t_2012_m

```

Mosaic plot shows which age group gave the biggest contribution in the observed difference in opinions:

```{r}
mosaicplot(cont_t_2012_m, shade = TRUE, las=2,
           main = "age_group")
```

The biggest decline from expected counts on this matter is for age group > 70 - there is much less people of that age suggest a woman have right to decide comparing to the situation when the distributed across age groups follows the general distriburion in population.

#### 3.3. Gender and opinion on abortions

Distribution of Yes and No answers for two genders:

```{r}
ggplot(abor_age_2012) +
  aes(x = sex, fill = abany) +
  geom_bar(position = "fill")
```

* * *

### Part 4: Inference

#### 4.1. Education level group and opinion on abortions

We compare two categorical variables, one of which with more than two levels, thus we can only use the Chi squared independence test, which is used when dealing with counts and investigating how far the observed counts are from the expected counts.

Conditions for the Chi squared independence test:

1. Independence: random sampling is provided by the data collection methodology, the data is from less of 10% of US adult population and one person contributed only.

2. Sample size: each particular scenario have more than 5 expected cases.

H0: independent
HA: dependent

```{r}
#===for 2012 ==============
chisq_edu_2012 <- chisq.test(cont_t_edu_2012)
chisq_edu_2012
```

As p-value << threshold value of 0.05 the H0 must be rejected.

```{r}
chisq_edu_2012$observed
round(chisq_edu_2012$expected,2)
round(chisq_edu_2012$residuals, 3)
```

The residuals graphs presents the deviations from what would be expected based on sampling variation. For a given cell, the size of the circle is proportional to the amount of the cell contribution:

```{r}
corrplot(chisq_edu_2012$residuals, is.cor = FALSE)
```

Let's also check out the contibution of each cell in percentage (%):

```{r}
contrib_edu <- 100*chisq_edu_2012$residuals^2/chisq_edu_2012$statistic
round(contrib_edu, 3)
```

#### 4.2. Age group and opinion on abortions 

We have the same structure of variables of interest with 4.1. Conditions for the Chi squared independence test are also valid.

H0: independent
HA: dependent

```{r}
chisq_2012 <- chisq.test(cont_t_2012)
chisq_2012 
```

Significance level obtained is less than the threshold value thus the H0 must be rejected.

```{r}
chisq_2012$observed
round(chisq_2012$expected,2)
round(chisq_2012$residuals, 3)
```

The residuals graphs presents the deviations from what would be expected based on sampling variation. For a given cell, the size of the circle is proportional to the amount of the cell contribution:
  
```{r}
corrplot(chisq_2012$residuals, is.cor = FALSE)
```

The contibution of each cell in percentage (%):

```{r}
contrib <- 100*chisq_2012$residuals^2/chisq_2012$statistic
round(contrib, 3)
corrplot(contrib, is.cor = FALSE)
```

The relative contribution of each cell to the total Chi-square score gives some indication of the nature of the dependency between rows and columns of the contingency table.

From the image above, it can be seen that the most contributing cells to the Chi-square scores are 45-49/Yes (12.75%), 45-49/No (10.21%), >70/Yes (22.58%), >70/No (18.10%).

These cells contribute about 63.64% to the total Chi-square score and thus account for most of the difference between expected and observed values.

#### 4.3. Gender and opinion on abortions

As two proportions to be compared to answer this research question, hypothesis test is to be used:
H0: p_male - p_female = 0
HA: p_male - p_female != 0

Conditions of applicability:
1. independence - random sample and < 10% of population
2. for males and females: numbers of successes and failures are all > 10
Hence p_male - p_female ~ N(mean = 0, SE = sqrt(p_pool * (1 - p_pool) * (1/n_male + 1/n_female)))

```{r}
table(abor_age_2012$sex)
cont_t_sex_2012 <- table(abor_age_2012$sex,abor_age_2012$abany)
cont_t_sex_2012

total_successes <- sum(cont_t_sex_2012[,1])
total_n <- sum(cont_t_sex_2012)

p_pool <- total_successes / total_n

n_male <- sum(cont_t_sex_2012[1,])
n_female <- sum(cont_t_sex_2012[2,])

SE <- sqrt(p_pool * (1 - p_pool) * (1/n_male + 1/n_female))

# point estimate p_hut_male - p_hut_female 
p_hut_male <- cont_t_sex_2012[1,1] / n_male
p_hut_female <- cont_t_sex_2012[2,1] / n_female

point_estimate <- p_hut_male - p_hut_female
Z <- (point_estimate - 0) / SE
Z
pnorm(Z)
```

There is no sufficient evidence to reject H0 - there is no statistically significant difference in opinions on abortion topic between males and females.

```{r}

inference(y = abany, x = sex, data = abor_age_2012, statistic = "proportion", type = "ht", null = 0, success="Yes", alternative ="greater", method = "theoretical")

```

* * *

### Part 5: Conclusion

Respondent's degree and respondent's age group are dependent on their opinion on abortions. Respondents with higher level of education and respondents in age group 35 - 49 years old answered Yes more frequently.

### References

[1] https://en.wikipedia.org/wiki/General_Social_Survey
